---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit the README.Rmd file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab report #3 - instructions

Follow the instructions posted at <https://ds202-at-isu.github.io/labs.html> for the lab assignment. The work is meant to be finished during the lab time, but you have time until Monday evening to polish things.

Include your answers in this document (Rmd file). Make sure that it knits properly (into the md file). Upload both the Rmd and the md file to your repository.

All submissions to the github repo will be automatically uploaded for grading once the due date is passed. Submit a link to your repository on Canvas (only one submission per team) to signal to the instructors that you are done with your submission.

# Lab 3: Avenger's Peril

## As a team

Extract from the data below two data sets in long form `deaths` and `returns`

Get the data into a format where the five columns for Death[1-5] are replaced by two columns: Time, and Death. Time should be a number between 1 and 5 (look into the function `parse_number`); Death is a categorical variables with values "yes", "no" and "". Call the resulting data set `deaths`.

```{r}
av <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/avengers/avengers.csv", stringsAsFactors = FALSE)
head(av)
```

```{r}
# View column names
colnames(av)

```

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(readr)

# Convert Death columns into long format and ensure correct categories
deaths <- av %>%
  select(Name.Alias, starts_with("Death")) %>%
  pivot_longer(cols = starts_with("Death"), 
               names_to = "Time", 
               values_to = "Death") %>%
  mutate(Time = readr::parse_number(Time),  # Convert Time to numeric
         Death = case_when(
           Death == "YES" ~ "yes",
           Death == "NO" ~ "no",
           TRUE ~ ""  # Keep empty values as they are
         )) %>%
  arrange(Name.Alias, Time)  # Sort the data for clarity

# View the first few rows to check the output
head(deaths)

```

```{r}
# Check structure of deaths dataset
str(deaths)

# View unique values in Death column
unique(deaths$Death)

```

Similarly, deal with the returns of characters.

```{r}
# Convert Return columns into long format and ensure correct categories
returns <- av %>%
  select(Name.Alias, starts_with("Return")) %>%
  pivot_longer(cols = starts_with("Return"), 
               names_to = "Time", 
               values_to = "Return") %>%
  mutate(Time = readr::parse_number(Time),  # Convert Time to numeric
         Return = case_when(
           Return == "YES" ~ "yes",
           Return == "NO" ~ "no",
           TRUE ~ ""  # Keep empty values as they are
         )) %>%
  arrange(Name.Alias, Time)  # Sort for clarity

# View the first few rows to check the output
head(returns)

```

Based on these datasets calculate the average number of deaths an Avenger suffers.

```{r}
# Calculate the average number of deaths per Avenger
avg_deaths <- deaths %>%
  filter(Death == "yes") %>%
  group_by(Name.Alias) %>%
  summarise(total_deaths = n()) %>%
  summarise(avg_deaths = mean(total_deaths))

avg_deaths

```

## Individually

For each team member, copy this part of the report.

Each team member picks one of the statements in the FiveThirtyEight [analysis](https://fivethirtyeight.com/features/avengers-death-comics-age-of-ultron/) and fact checks it based on the data. Use dplyr functionality whenever possible.

### FiveThirtyEight Statement (Selim's Individual Work)

> "Out of 173 listed Avengers, my analysis found that 69 had died at least one time after they joined the team. That’s about 40 percent of all people who have ever signed on to the team."

```{r Selim Statistics 1}
# How many Avengers have returned at least once?
avengers_returned <- returns %>%
  filter(Return == "yes") %>%
  summarise(unique_returns = n_distinct(Name.Alias))

# What percentage of Avengers have returned?
total_avengers <- n_distinct(deaths$Name.Alias)

return_percentage <- (avengers_returned$unique_returns / total_avengers) * 100

# Average number of returns per Avenger
avg_returns <- returns %>%
  filter(Return == "yes") %>%
  group_by(Name.Alias) %>%
  summarise(total_returns = n()) %>%
  summarise(avg_returns = mean(total_returns))

# Print results
avengers_returned
return_percentage
avg_returns
```

This means that on average, an Avenger comes back to life about 1.27 times in the comics.

```{r Selim Statistics 2}
avengers_died <- deaths %>%
  filter(Death == "yes") %>%
  summarise(unique_deaths = n_distinct(Name.Alias))

death_percentage <- (avengers_died$unique_deaths / total_avengers) * 100

death_percentage

```

Conclusion (FiveThirtyEight's vs Selims Results) Based on the dataset, the percentage of Avengers who have died at least once is 39.26%, which is very close to FiveThirtyEight's claim of 40%. Their claim appears to be accurate.

### FiveThirtyEight Statement (Kush's Individual Work)

> “The Avengers roster is overwhelmingly male — only 8 percent of all the people who have been Avengers are women.”

```{r your-statement}
# Total number of unique Avengers

total_avengers <- av %>%
  filter(!is.na(Name.Alias)) %>%
  summarise(n = n_distinct(Name.Alias)) %>%
  pull(n)

# Total number of unique FEMALE Avengers
female_avengers <- av %>%
  filter(Gender == "FEMALE") %>%
  summarise(n = n_distinct(Name.Alias)) %>%
  pull(n)

# Calculate percentage
female_percent <- (female_avengers / total_avengers) * 100
female_percent
```

### FiveThirtyEight Statement (Ash’s Individual Work)

> “Many Avengers have died multiple times — some even five times!”

```{r}
# Count how many times each Avenger has died
death_counts <- deaths %>%
  filter(Death == "yes") %>%
  group_by(Name.Alias) %>%
  summarise(num_deaths = n()) %>%
  arrange(desc(num_deaths))

# Count Avengers who died more than once
multi_deaths <- death_counts %>%
  filter(num_deaths > 1) %>%
  summarise(count = n())

# Count Avengers who died exactly five times
five_deaths <- death_counts %>%
  filter(num_deaths == 5) %>%
  summarise(count = n())

multi_deaths
five_deaths
```

### Conclusion:

According to my analysis (Kush), approximately `r round(female_percent, 2)`% of Avengers are female — significantly higher than the 8% reported by FiveThirtyEight. This means the original statement underestimates the representation of women among Avengers in this dataset. The claim appears inaccurate.

According to my analysis (Ash), several Avengers have died multiple times — with `r five_deaths$count` character(s) dying exactly five times and `r multi_deaths$count` dying more than once. This confirms the FiveThirtyEight statement that “many Avengers have died multiple times — some even five times.” The claim appears accurate based on the dataset.

### FiveThirtyEight Statement

> Quote the statement you are planning to fact-check.

### Include the code

Make sure to include the code to derive the (numeric) fact for the statement

### Include your answer

Include at least one sentence discussing the result of your fact-checking endeavor.

Upload your changes to the repository. Discuss and refine answers as a team.
